---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# jlview — Zero-Copy Julia↔R Arrays via ALTREP

<!-- badges: start -->
[![R-CMD-check](https://github.com/tanaylab/jlview/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/tanaylab/jlview/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

jlview provides zero-copy R views of Julia-owned arrays using R's ALTREP (Alternative Representations) framework. Instead of copying data between Julia and R, jlview returns lightweight R vectors that point directly into Julia's memory.

| | Latency | R Memory |
|---|---|---|
| **jlview (zero-copy)** | 0.02 ms | 0.1 MB |
| **copy (collect)** | 0.44 ms | 76.3 MB |
| **Improvement** | **22.6×** | **99.9% savings** |

*Benchmark: 10K × 1K named Float64 matrix*

## Installation

```r
# install.packages("remotes")
remotes::install_github("tanaylab/jlview")
```

### Requirements

- R ≥ 4.0
- [Julia](https://julialang.org/) ≥ 1.6
- [JuliaCall](https://github.com/JuliaInterop/JuliaCall) R package

## Usage

```{r example, eval=FALSE}
library(jlview)
JuliaCall::julia_setup()

# Create a Julia array
JuliaCall::julia_command("x = randn(10000, 1000)")

# Zero-copy view — R sees Julia's memory directly
m <- jlview(JuliaCall::julia_eval("x"))
dim(m)    # [1] 10000  1000
sum(m)    # works natively, no data copied

# Named arrays are supported
JuliaCall::julia_command("using NamedArrays")
JuliaCall::julia_command("named = NamedArray(randn(3), [\"a\", \"b\", \"c\"])")
v <- jlview_named_vector(JuliaCall::julia_eval("named"))
v["a"]    # access by name, still zero-copy

# Sparse matrices (zero-copy values, shifted indices)
JuliaCall::julia_command("using SparseArrays")
JuliaCall::julia_command("sp = sprand(1000, 500, 0.01)")
s <- jlview_sparse(JuliaCall::julia_eval("sp"))
class(s)  # "dgCMatrix"

# Explicit release to free Julia memory early
jlview_release(m)
```

## Key Features

- **Zero-copy dense arrays** — Float64, Int32 map directly to R's REALSXP, INTSXP
- **Type conversion** — Float32, Int64, Int16 are converted once in Julia, then zero-copy to R
- **Named arrays** — NamedArray row/column names attached atomically without triggering copy
- **Sparse matrices** — `dgCMatrix` with zero-copy values (`nzval`) and shifted indices
- **Copy-on-write** — R's standard COW semantics: reads are zero-copy, writes trigger materialization
- **GC safety** — Julia arrays are pinned while R holds references; three-layer defense (pinning dict, memory pressure tracking, explicit release)
- **Fork safety** — Safe with `parallel::mclapply()` (PID-guarded finalizers)
- **Serialization** — `saveRDS()`/`readRDS()` work correctly (materializes on save)

## Supported Types

| Julia type | R type | Method |
|---|---|---|
| `Array{Float64}` | `numeric` | Direct zero-copy |
| `Array{Int32}` | `integer` | Direct zero-copy |
| `Array{Float32}` | `numeric` | Convert in Julia, then zero-copy |
| `Array{Int64}` | `numeric` | Convert in Julia, then zero-copy |
| `Array{Int16}` | `integer` | Convert in Julia, then zero-copy |
| `Array{UInt8}` | `raw` | ALTREP RAWSXP zero-copy |
| `Array{Bool}` | `logical` | Copy (layout incompatible) |
| `String[]` | `character` | Copy (layout incompatible) |

## Integration with dafr

jlview is used by [dafr](https://github.com/tanaylab/dafr) to provide zero-copy access to [DataAxesFormats.jl](https://github.com/tanaylab/DataAxesFormats.jl) data. When using dafr with jlview installed, `get_vector()`, `get_matrix()`, and sparse matrix operations automatically use zero-copy where possible.

## How It Works

jlview uses R's [ALTREP](https://svn.r-project.org/R/branches/ALTREP/ALTREP.html) framework to create R vectors backed by Julia memory:

1. **Pin** — The Julia array is stored in a global dictionary, preventing Julia's GC from collecting it
2. **Wrap** — An ALTREP R vector is created whose `Dataptr` returns Julia's raw data pointer
3. **Use** — R operations (sum, subsetting, etc.) read directly from Julia's memory
4. **Release** — When R garbage-collects the ALTREP object, a C finalizer calls Julia to unpin the array

The entire pin→ALTREP→finalizer path is implemented in C using Julia's C API (`jl_call1`), avoiding JuliaCall overhead and ensuring safety during R's garbage collection.
